---
title: "Assignment 1 - Language Development in ASD - part 3"
author: "Nanna Kildahl Mathiasen"
date: "August 10, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Welcome to the third exciting part of the Language Development in ASD exercise

In this exercise we will delve more in depth with different practices of model comparison and model selection, by first evaluating your models from last time, then learning how to cross-validate models and finally how to systematically compare models.

N.B. There are several datasets for this exercise, so pay attention to which one you are using!

1. The (training) dataset from last time (the awesome one you produced :-) ).
2. The (test) datasets on which you can test the models from last time:
* Demographic and clinical data: https://www.dropbox.com/s/ra99bdvm6fzay3g/demo_test.csv?dl=0
* Utterance Length data: https://www.dropbox.com/s/uxtqqzl18nwxowq/LU_test.csv?dl=0
* Word data: https://www.dropbox.com/s/1ces4hv8kh0stov/token_test.csv?dl=0

```{r}
getwd()
locpath= "/Users/nannakildahlmathiasen/OneDrive/3 Semester/Experimental Methods 3/Assignment 1"
setwd(locpath)
train_data = read.csv("finaldata.csv")
demo_data = read.csv("demo_test.csv")
lu_data = read.csv("LU_test.csv")
word_data = read.csv("token_test.csv")

#Omit the first column
train_data = train_data[,-1]



library(lmerTest)
library(MuMIn)
library(modelr)
library(plyr)
library(stringr)
library(tidyverse)
library(merTools)
library(caret)
library(Metrics)
library(ggplot2)

```

### Exercise 1) Testing model performance

How did your models from last time perform? In this exercise you have to compare the results on the training data () and on the test data. Report both of them. Compare them. Discuss why they are different.

- recreate the models you chose last time (just write the code again and apply it to Assignment2TrainData1.csv)
- calculate performance of the model on the training data: root mean square error is a good measure. (Tip: google the functions rmse() and predict() )
- create the test dataset (apply the code from assignment 1 part 1 to clean up the 3 test datasets)
- test the performance of the models on the test data (Tips: time to reuse "predict()")
- optional: predictions are never certain, can you identify the uncertainty of the predictions? (e.g. google predictinterval())

```{r}
train_model = lmer(CHI_MLU ~ Visit + MOT_MLU + verbalIQ + (1 + Visit|ID), train_data, REML = FALSE)
summary(train_model) 

# To find the root mean square error
r.squaredGLMM(train_model)


#Calculating performance of the model on the training data:
modelr::rmse(train_model, train_data)


#clean the 3 datasets:
names(demo_data)[names(demo_data)=="Child.ID"] <- "ID"
names(lu_data)[names(lu_data)=="SUBJ"] <- "ID"
names(word_data)[names(word_data)=="SUBJ"] <- "ID"

names(demo_data)[names(demo_data)=="Visit"] <- "Visit"
names(lu_data)[names(lu_data)=="VISIT"] <- "Visit"
names(word_data)[names(word_data)=="VISIT"] <- "Visit"

#To homogeneize the way "visit" is reported
library(stringr)

demo_data$Visit = str_extract(demo_data$Visit, "\\d")
lu_data$Visit = str_extract(lu_data$Visit, "\\d")
word_data$Visit = str_extract(word_data$Visit, "\\d")

#To remove the "." at the end of names
demo_data$ID = gsub("\\.","",demo_data$ID)
lu_data$ID = gsub("\\.","",lu_data$ID)
word_data$ID = gsub("\\.","",word_data$ID)

#To get the variables we need: Child.ID, Visit, Ethnicity, Diagnosis, Gender, Age, ADOS,  MullenRaw, ExpressiveLangRaw, MOT_MLU, MOT_LUstd, CHI_MLU, CHI_LUstd, types_MOT, types_CHI, tokens_MOT, tokens_CHI
library(tidyverse)
demo_data_sub = dplyr::select(demo_data, ID, Visit, Ethnicity, Diagnosis, Gender, Age, ADOS, MullenRaw, ExpressiveLangRaw)
lu_data_sub = dplyr::select(lu_data, ID, Visit, MOT_MLU, MOT_LUstd, CHI_MLU, CHI_LUstd)
word_data_sub = dplyr::select(word_data, ID, Visit, types_MOT, types_CHI, tokens_MOT, tokens_CHI)


names(demo_data_sub)[names(demo_data_sub)=="MullenRaw"] <- "nonverbalIQ"
names(demo_data_sub)[names(demo_data_sub)=="ExpressiveLangRaw"] <- "verbalIQ"

#To merge the three subsets
sub_data_demolu = merge(demo_data_sub, lu_data_sub)
test_data = merge(sub_data_demolu, word_data_sub)

test_data_sub = subset(test_data, Visit == "1")
test_data_sub = dplyr::select(test_data_sub, ID, ADOS, nonverbalIQ, verbalIQ)

#To omit the variables ADOS, nonVerbalIQ (MullenRaw) and verbalIQ (ExpressiveLangRaw)
test_data = test_data[,-7:-9]
final_test_data = merge(test_data, test_data_sub)


final_test_data$ID = as.numeric(as.factor(as.character(final_test_data$ID)))
final_test_data$Visit = as.numeric(as.character(final_test_data$Visit))

final_test_data$Gender = as.factor(final_test_data$Gender)
final_test_data <- transform(final_test_data,
          Gender=revalue(Gender,c("1"="M", "2"="F")))

final_test_data <- transform(final_test_data,
          Diagnosis=revalue(Diagnosis,c("A"="ASD", "B"="TD")))


# To test the model on the test dataset
Metrics::rmse(final_test_data$CHI_MLU, predict(train_model, final_test_data, allow.new.levels = T))

# To test the model on the train dataset
Metrics::rmse(train_data$CHI_MLU, predict(train_model, train_data, allow.new.levels = T))

# To get prediction interval for each participant for each visit 
predictInterval(merMod = train_model, newdata = final_test_data, level = 0.95, n.sims = 1000, stat = "median", type="linear.prediction",include.resid.var = TRUE)
```

### Exercise 2) Model Selection via Cross-validation (N.B: ChildMLU!)

One way to reduce bad surprises when testing a model on new data is to train the model via cross-validation. 

In this exercise you have to use cross-validation to calculate the predictive error of your models and use this predictive error to select the best possible model.

- Create the basic model of ChildMLU as a function of Time and Diagnosis (don't forget the random effects!).
- Make a cross-validated version of the model. (Tips: google the function "createFolds";  loop through each fold, train a model on the other folds and test it on the fold)
- Report the results and comment on them.

- Now try to find the best possible predictive model of ChildMLU, that is, the one that produces the best cross-validated results.

```{r}
library(caret)
library(Metrics)

#Creating 5 folds of the train data, where it divides by ID. 
train_data$ID = as.factor(train_data$ID)
folds = createFolds(unique(train_data$ID),4)

train_RMSE = NULL
test_RMSE = NULL
n = 1

for (fold in folds) {
  #train2 is everything but not fold, in first loop fold 1, in second loop fold 2. ! means all but not the following
  train2 = subset(train_data, ! (ID %in% fold))
  #test2 is every other fold than the fold in train2
  test2 = subset(train_data, (ID %in% fold))
  #The model we want to test
  train_model_2 = lmer(CHI_MLU ~ Visit + Diagnosis + (1 + Visit|ID), train2, REML = FALSE)
  #Test the model on train data, seeing how well the model predicts train data
  train_RMSE[n] = Metrics::rmse(train2$CHI_MLU, predict(train_model_2, train2))
  #Test the model on test data, seeing how well the model predicts test data - which is not a part of the model
  pred = predict(train_model_2, test2, allow.new.levels = TRUE)
  test_RMSE[n] = Metrics::rmse(test2$CHI_MLU, pred)
  n = n+1
}
#Reporting the mean of how well the model predicts. 
mean(test_RMSE)
mean(train_RMSE)
#The model is not a good model, because the model has much more error when predicting from the test data (rmse = 0.81) compared to the train data (rsme = 0.34). 


#Testing the model we constructed last time and see if it predicts better:
train_RMSE3 = NULL
test_RMSE3 = NULL
n = 1
for (fold in folds) {
  #train3 is everything but not fold, in first loop fold 1, in second loop fold 2. ! means all but not the following
  train3 = subset(train_data, ! (ID %in% fold))
  #test3 is every other fold than the fold in train3
  test3 = subset(train_data, (ID %in% fold))
  #The better model we want to test
  train_model_3 = lmer(CHI_MLU ~ Visit + MOT_MLU + verbalIQ + (1 + Visit|ID), train3, REML = FALSE)
  #Test the model on train data, seeing how well the model predicts train data
  train_RMSE3[n] = Metrics::rmse(train3$CHI_MLU, predict(train_model_3, train3))
  #Test the model on test data, seeing how well the model predicts test data - which is not a part of the model
  pred = predict(train_model_3, test3, allow.new.levels = TRUE)
  test_RMSE3[n] = Metrics::rmse(test3$CHI_MLU, pred)
  n = n+1
}
#Reporting the mean of how well the model predicts. 
mean(test_RMSE3)
mean(train_RMSE3)
#The model is a better model because the error difference is smaller, test data (rmse = 0.61) and train data (rmse = 0.34). 

#Testing a new model:
train_RMSE4 = NULL
test_RMSE4 = NULL
n = 1
for (fold in folds) {
  #train3 is everything but not fold, in first loop fold 1, in second loop fold 2. ! means all but not the following
  train4 = subset(train_data, ! (ID %in% fold))
  #test3 is every other fold than the fold in train3
  test4 = subset(train_data, (ID %in% fold))
  #The better model we want to test
  train_model_4 = lmer(CHI_MLU ~ Visit*Diagnosis + MOT_MLU + verbalIQ + (1 + Visit|ID), train4, REML = FALSE)
  #Test the model on train data, seeing how well the model predicts train data
  train_RMSE4[n] = Metrics::rmse(train4$CHI_MLU, predict(train_model_4, train4))
  #Test the model on test data, seeing how well the model predicts test data - which is not a part of the model
  pred = predict(train_model_4, test4, allow.new.levels = TRUE)
  test_RMSE4[n] = Metrics::rmse(test4$CHI_MLU, pred)
  n = n+1
}
#Reporting the mean of how well the model predicts. 
mean(test_RMSE4)
mean(train_RMSE4)
#The model is a better model because the error difference is smaller, test data (rmse = 0.56) and train data (rmse = 0.33). 

```

- Bonus Question 1: How would you go comparing the performance of the basic model and the cross-validated model on the testing set?
- Bonus Question 2: What is the effect of changing the number of folds? Can you plot RMSE as a function of number of folds?
- Bonus Question 3: compare the cross-validated predictive error against the actual predictive error on the test data

### Exercise 3) Assessing the single child

Let's get to business. This new kiddo - Bernie - has entered your clinic. This child has to be assessed according to his group's average and his expected development.

Bernie is one of the six kids in the test dataset, so make sure to extract that child alone for the following analysis.

You want to evaluate:

- how does the child fare in ChildMLU compared to the average TD child at each visit? Define the distance in terms of absolute difference between this Child and the average TD.
(Tip: recreate the equation of the model: Y=Intercept+BetaX1+BetaX2, etc; input the average of the TD group  for each parameter in the model as X1, X2, etc.).

- how does the child fare compared to the model predictions at Visit 6? Is the child below or above expectations? (tip: use the predict() function on Bernie's data only and compare the prediction with the actual performance of the child)


```{r}
#Making a subset with only Bernie
bernie = merge(sub_data_demolu, word_data_sub)

bernie_sub = subset(bernie, Visit == "1")
bernie_sub = dplyr::select(bernie_sub, ID, ADOS, nonverbalIQ, verbalIQ)
bernie = bernie[,-7:-9]
bernie1 = merge(bernie, bernie_sub, by = "ID")
bernie1 = subset(bernie1, ID == "Bernie",)
bernie1$ID = as.numeric(as.factor(as.character(bernie1$ID)))
bernie1$Visit = as.numeric(as.character(bernie1$Visit))

#Transform gender and diagnosis variable
bernie1$Gender = as.factor(bernie1$Gender)
bernie1 <- transform(bernie1,
          Gender=revalue(Gender,c("1"="M", "2"="F")))
bernie1 <- transform(bernie1,
          Diagnosis=revalue(Diagnosis,c("A"="ASD", "B"="TD"))) 


#Making a dataframe with only data from TD's
TDtrain = train_data[train_data$Diagnosis == "TD",]
#Making a model for the TD children
TDmodel = lmer(CHI_MLU ~ 1 + Visit + MOT_MLU + verbalIQ + (1 + Visit|ID), TDtrain, REML = FALSE)
summary(TDmodel)

# To compare Bernie to the average TD
# Equation: -1.69 + 0.285*x (VISIT) + 0.4896*y (MOT_MLU) + 0.0456*z (verbalIQ)
myfunction = function(x, y, z) {
  -1.69 + 0.285*x + 0.4896*y + 0.0456*z
}

mean(TDtrain$MOT_MLU[TDtrain$Visit == 1])
mean(TDtrain$MOT_MLU[TDtrain$Visit == 2])
mean(TDtrain$MOT_MLU[TDtrain$Visit == 3])
mean(TDtrain$MOT_MLU[TDtrain$Visit == 4])
mean(TDtrain$MOT_MLU[TDtrain$Visit == 5])
mean(TDtrain$MOT_MLU[TDtrain$Visit == 6])
MEAN_verbalIQ = mean(TDtrain$verbalIQ)

# Visit 1
TD_visit1=myfunction(1, 3.725789, 20.15)
dif_visit1 = bernie1$CHI_MLU[bernie1$Visit == 1] - TD_visit1
dif_visit1

# Visit 2
TD_visit2=myfunction(2, 3.974927 ,20.15)
dif_visit2 = bernie1$CHI_MLU[bernie1$Visit == 2] - TD_visit2
dif_visit2

# Visit 3
TD_visit3=myfunction(3, 4.147266 ,20.15)
dif_visit3 = bernie1$CHI_MLU[bernie1$Visit == 3] - TD_visit3
dif_visit3

# Visit 4
TD_visit4=myfunction(4, 4.289469 ,20.15)
dif_visit4 = bernie1$CHI_MLU[bernie1$Visit == 4] - TD_visit4
dif_visit4

# Visit 5
TD_visit5=myfunction(5, 4.407844 ,20.15)
dif_visit5 = bernie1$CHI_MLU[bernie1$Visit == 5] - TD_visit5
dif_visit5

# Visit 6
TD_visit6=myfunction(6, 4.39503 ,20.15)
dif_visit6 = bernie1$CHI_MLU[bernie1$Visit == 6] - TD_visit6
dif_visit6

# To create a dataframe 
Visits = c(1,2,3,4,5,6)
Category = c("TD", "TD","TD","TD","TD","TD", "Bernie","Bernie","Bernie","Bernie","Bernie","Bernie")
MLU = c(TD_visit1, TD_visit2, TD_visit3, TD_visit4, TD_visit5, TD_visit6, bernie1$CHI_MLU)

Bernie_comp=data.frame(Visits, Category, MLU)

# To create a plot
ggplot(Bernie_comp, aes(x = Visits, y= MLU, colour = Category)) + geom_smooth() + geom_point()
# From the plot it seems that Bernie performs better than the average TD (except Visit 2)

#Predicting the expected CHI_MLU for Bernie
bernie_pred = predict(TDmodel, bernie1, allow.new.levels = TRUE)
#Calculating the distance between the actual and the predicted value of CHI_MLU for Bernie
distance = bernie1$CHI_MLU - bernie_pred
distance
#Bernie develops better than TD children except for visit 2

# Using the predict function to compare VISIT 6
Bernie_6 = bernie1[6,] # To subset visit 6
Bernie_pred=predict(TDmodel, Bernie_6, allow.new.levels = T) 
Bernie_6$CHI_MLU - Bernie_pred # To get the difference between predicted and acutual value

```

### OPTIONAL: Exercise 4) Model Selection via Information Criteria
Another way to reduce the bad surprises when testing a model on new data is to pay close attention to the relative information criteria between the models you are comparing. Let's learn how to do that!

Re-create a selection of possible models explaining ChildMLU (the ones you tested for exercise 2, but now trained on the full dataset and not cross-validated).

Then try to find the best possible predictive model of ChildMLU, that is, the one that produces the lowest information criterion.

- Bonus question for the optional exercise: are information criteria correlated with cross-validated RMSE? That is, if you take AIC for Model 1, Model 2 and Model 3, do they co-vary with their cross-validated RMSE?

### OPTIONAL: Exercise 5): Using Lasso for model selection
Welcome to the last secret exercise. If you have already solved the previous exercises, and still there's not enough for you, you can expand your expertise by learning about penalizations. Check out this tutorial: http://machinelearningmastery.com/penalized-regression-in-r/ and make sure to google what penalization is, with a focus on L1 and L2-norms. Then try them on your data!


